Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2023-02-06 src/docs_paper/paper.Rmd
Good writing progress: mostly saying what an LGM and ELGM are

---
title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via Template Model Builder \texttt{TMB}. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides substantially more accurate inferences than default \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler but substantially faster. By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior, enabling inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at the level of geographic resolution at which heath systems are planned and delivered.
Producing these estimates is challenging because all available data sources have shortcomings which must be overcome.
Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they occur only every five years or so, with limited sample size at the district level.
Other data sources, such as routine health surveillance of antenatal care clinics, are more real-time but based on limited or non-representative samples of the population.
To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
Simultaneous modelling of the data mitigates the limitations of any single source, increases statistical power, and prompts investigation into any conflicting information.
Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
The creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.

The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments, which for the Naomi model, we use for the high-dimensional latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
This has motivated us to look for a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.

@@

The remainder of this paper is organised as follows.
Section \ref{sec:naomi} describes a simplified version of the Naomi model that we consider in this paper.
Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference using simplified INLA and AGHQ.
As a case-study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how we anticipate our method might be useful for other models, and directions for future research.

@@

@eaton2021naomi specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence among pregnant women from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
The full model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version of the model which is defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
Below we provide an overview of the three components in the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

@@

Consider a country in sub-Saharan Africa where a household survey has taken place during at time $T_1$.
Let $x$ index district, $a$ five-year age band, and $s$ sex and.
For ease of notation, let $i$ index the finest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be ART coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale
$$
\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
$$
where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
$$
Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
$$
where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$.

@@

@@

Consider a latent Gaussian model (LGM) of the form
\begin{alignat*}{2}
&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \\
&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \\
&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta),
\end{alignat*}
where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
The joint posterior of $(\x, \btheta)$ is given by
\begin{equation*}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
\end{equation*}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
\begin{align}
  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}

@@

1. Calculate the mode, Hessian at the mode, and lower Cholesky
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
\end{align*}
of the Laplace approximation
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $m = \dim(\btheta)$, and $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
We typically set $k = 3$ such that there are $3^m$ nodes in total.
3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
$$
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
$$
4. For $i = 1, \ldots, n$ generate a set of nodes $\mathcal{X}_i$ via a Gauss-Hermite quadrature rule in one-dimension adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$ to evaluate each latent field marginal Laplace approximation at.
5. For $x_i \in \mathcal{X}_i$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ may be initialised at $\hat \x(\btheta)_{-i}$.
6. For $x_i \in \mathcal{X}_i$ calculate
$$
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
$$
where
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
$$
and
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
7. Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior summaries can be obtained.

>>>>

---
title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A,D
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. Inference for Naomi is currently conducted using an empirical Bayes Gaussian approximation via the \texttt{TMB} \textsc{R} package. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides substantially more accurate inferences than \texttt{TMB}, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler, but substantially faster. By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior, enabling inference via integrated nested Laplace approximations for a larger class of models than was previously possible.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

Mounting an effective public health response to the HIV epidemic requires accurate, timely HIV indicator estimates at the level of geographic resolution at which heath systems are planned and delivered.
Producing these estimates is challenging because all available data sources have shortcomings which must be overcome.
Nationally-representative household surveys provide the most statistically reliable data, but due to their high cost, in most countries they occur only every five years or so, with limited sample size at the district level.
Other data sources, such as routine health surveillance of antenatal care (ANC) clinics, are more real-time but based on limited or non-representative samples of the population.
To meet these challenges, the Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV indicators at a district-level.
Simultaneous modelling of the data mitigates the limitations of any single source, increases statistical power, and prompts investigation into any conflicting information.
Software (\url{https://naomi.unaids.org}) has been developed for Naomi, allowing countries to input their data and interactively generate estimates in a yearly process supported by UNAIDS.
The creation of estimates by country teams, rather than external agencies, is a noteworthy feature of the HIV response.
Drawing on expertise closest to the data being modelled improves the accuracy of the process, as well as strengthening trust and ownership of the estimates.

The use case requirements for the model, combined with its relative complexity, present a difficult Bayesian inference problem.
Any inferential strategy must be fast enough for interactive review and iteration of modelling results, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes (EB) approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` has recently been gaining popularity more broadly in spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments, which for the Naomi model, we use for the high-dimensional latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.
This has motivated us to look for a more accurate approach, which is flexible enough to be compatible with the model and fast enough to be run in production by country teams.

@@

Other work aiming to extend the scope of the INLA method includes the `inlabru` \textsc{R} package of @bachl2019inlabru, and the INLA within MCMC approach of @gomez2018markov, both of which leverage the `R-INLA` \textsc{R} package [@martins2013bayesian].
To approximate non-linear predictors, `inlabru` makes iterated use of `R-INLA` using linearisation, extending the scope of INLA to LGMs with some suitably small amount of non-linearity.
<!-- See https://inlabru-org.github.io/inlabru/articles/method.html -->
<!-- Want to say in particular what the class of models that `inlabru` allows you to fit are -->
INLA within MCMC is suitable for models which are LGMs, though only conditional on some subset of the parameters being fixed.
For these models, Metropolis-Hastings style algorithms be defined to update the fixed parameters, with acceptance probabilities calculated using by repeat calls to `R-INLA`.
It would or would not be possible to fit the Naomi model in these frameworks.

The remainder of this paper is organised as follows.
Section \ref{sec:naomi} describes a simplified version of the Naomi model that we consider in this paper.
Section \ref{sec:elgm} describes how the Naomi model falls within the ELGM framework.
Section \ref{sec:fastinferencemethods} outlines our approach to fast, accurate Bayesian inference for ELGMs using simplified INLA and AGHQ.
As a case-study, we compare the accuracy of our inference method to `TMB` and `tmbstan` for the simplified Naomi model fit to data from Malawi, in Section \ref{sec:naomi}.
We also demonstrate a Bayesian workflow, illustrating the applicability of these tools in a deterministic inference setting.
Finally, in Section \ref{sec:conclusions} we discuss our conclusions, how we anticipate our method might be useful for other models, and directions for future research.

@@

@eaton2021naomi specify a joint model linking three small-area estimation models.
The full model is defined over three time points: $T_1$ the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version defined only at $T_1$ omitting nowcasting of $T_2$ and temporal projection to $T_3$.
Below we provide an overview of the simplified model, highlighting the aspects which make it a challenge for existing inferential approaches.
A more complete mathematical description of the simplified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

@@

Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place during at time $T_1$.
Let $x$ index district, $a$ five-year age band, and $s$ sex.
For ease of notation, let $i$ index the finest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale
$$
\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
$$
where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
$$
Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
$$
where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

@@

People living with HIV may choose to access ART services outside of the district that they reside in.
Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent $x \sim x'$, and model $\gamma_{x, x'}$ using a multinomial logistic regression model where $\tilde \gamma_{x, x'}$ are the log-odds.

@@

# Extended Latent Gaussian models\label{sec:elgm}

Latent Gaussian models (LGMs) [@rue2009approximate] are of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $[n] = \{1, \ldots, n\}$.
The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
The vector $\btheta_1 \in \mathbb{R}^s$, with $s_1$ assumed small, are additional parameters of the likelihood.
The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.

Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ depends on, but may not be equal to, $n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this allows for the presence of more non-linearity in the model than is allowed in LGMs.
ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
\end{align*}

The Naomi model falls into the ELGM class as follows.

@@

The joint posterior of $(\x, \btheta)$ for an ELGM is given by
\begin{equation*}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
\end{equation*}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and hyperparameter $\theta_j$ given by
\begin{align}
  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i \in [N], \label{eq:inla1} \\
  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j \in [m]. \label{eq:inla2}
\end{align}

@@

1. Calculate the mode, Hessian at the mode, and lower Cholesky
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\text{LA}(\btheta, \y)}, \\
\mathbf{H} &= \frac{\partial^2}{\partial \btheta \partial \btheta^\top} - \log \tilde p_\text{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\mathbf{H}^{-1} &= \mathbf{L} \mathbf{L}^\top,
\end{align*}
of the Laplace approximation
$$
\tilde p_\text{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmin_\x - \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= \frac{\partial^2}{\partial x \partial x^\top} - \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega: \z \in \mathcal{Q}(m, k) \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\z) = \hat{\btheta} + \mathbf{L} \z$.
If possible $k \geq 3$ is prefered, though the number of grid points scales exponentially with choice of $k$.
3. Use this quadrature rule to calculate the normalising constant $\tilde p_{\text{AQ}}(\y)$ as follows
$$
\tilde p_{\text{AQ}}(\y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(\btheta(\z), \y) \omega(\z).
$$
4. For $i \in [n]$ generate $l$ nodes $\mathcal{X}_i$ via a Gauss-Hermite quadrature rule with one-dimension adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\mathbf{H}(\btheta)^{-1}]_i}$. A value of $l \geq 4$ is recommended to enable B-spline interpolation.
5. For $x_i \in \mathcal{X}_i$ and $\btheta \in \{ \btheta(\z) \}_{\z \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmin_{\x_{-i}} - \log p(\y, x_i, \x_{-i}, \btheta), \\
\mathbf{H}_{-i, -i}(x_i, \btheta) &= \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} - \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
6. For $x_i \in \mathcal{X}_i$ calculate
$$
\tilde p_\text{AQ}(x_i \, | \, \y) = \frac{\tilde p_\text{LA}(x_i, \y)}{p_{\text{AQ}}(\y)}.
$$
where
$$
\tilde p_\text{LA}(x_i, \y) = \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\text{LA}(x_i, \btheta(\z), \y) \omega(\z).
$$
and
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
7. Given $\{x_i, \tilde p_\text{AQ}(x_i \, | \, \y)\}_{x_i \in \mathcal{X}_i}$ create a spline interpolant to each posterior marginal on the log-scale, from which samples or relevant posterior marginal summaries may be obtained.
